---
title: "Using lgpr with ordinal response"
author: "Juho Timonen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
#  pdf_document:
#    toc: true     
vignette: >
  %\VignetteIndexEntry{Using lgpr with ordinal response}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#"
)
```

The ordinal observation model is currently not part of the master branch of `lgpr`, but you can install it with `devtools::install_github('jtimonen/lgpr', ref='ordinal')`.

```{r load}
library(ggplot2)
library(lgpr)
```

## Simulating ordinal data
We first simulate a process.
```{r simulate, fig.width=7.2, fig.height=4}
# Simulate data
set.seed(123)
cuts <- c(-1,0.5)
lc1 <- 'firebrick3'
lc2 <- "steelblue3"
simData <- simulate_data(N=6, t_data=c(10,20,30,40,50,60,70,80,90,100))
dat <- simData$data
p <- plot_data(dat) + 
  geom_hline(yintercept = cuts[1], color = lc1) +
  geom_hline(yintercept = cuts[2], color = lc2)
p <- p + ggtitle('Simulation: horizontal lines show cut points')
p
```

We discretize the response into values 1, 2 and 3 using cut points, i.e. thresholds, -1 (red line) and 0.5 (blue line).
```{r ord, fig.width=7.2, fig.height=4}
# Make response ordinal
y <- dat$y
y_ord <- rep(2, length(y))
y_ord[y < cuts[1]] <- 1
y_ord[y > cuts[2]] <- 3
dat$y <- y_ord
p <- plot_data(dat)
p + theme(panel.grid.minor = element_blank())
```

## Fitting a model with ordinal response
Now we fit an additive GP model $f^{(sum)} = f^{(1)}(id, age) + f^{(2)}(age)$ with ordinal observation model for response $y$.
```{r fit, fig.width=7.2, fig.height=4}
# Fit model
fit <- lgp(formula    = y ~ id + age, 
           data       = dat, 
           likelihood = 'ordinal',
           control    = list(adapt_delta=0.95),
           parallel   = TRUE,
           refresh    = 0)
```

*Note*: computing the proportion of noise is not yet implemented in a sensible way for ordinal responses, so do not trust the inferred relevance values.
```{r show, fig.width=7.2, fig.height=4}
print(fit)
```

## Visualising components and inferred cut points
We plot the average inferred function $f^{(sum)} = f^{(1)}(id, age) + f^{(2)}(age)$ and the posterior means of the cut points
```{r vis1, fig.width=7.2, fig.height=4}
# Get function components
ff <- get_function_components(fit)
f_sum <- ff[,,3]
f_sum_mean <- apply(f_sum, 2, mean)
df <- data.frame(dat$id, dat$age, f_sum_mean)
colnames(df) <- c("id", "age", "f_sum") 
plt <- ggplot(df, aes(x=age, group=id, y=f_sum)) + geom_line()

# Get inferred cutpoints
cut_points <- get_cutpoints(fit)
cmeans <- colMeans(cut_points)
cstds  <- apply(cut_points, 2, sd)
plt <- plt + geom_hline(yintercept=cmeans[1], color = lc1, alpha=1)
plt <- plt + geom_hline(yintercept=cmeans[2], color = lc2, alpha=11)
plt <- plt + ggtitle('Horizontal lines show inferred cut points')
plt
```

We can also visualize their variance
```{r vis2, fig.width=7.2, fig.height=5}
# Plot all samples of cut points
plt <- plt + geom_hline(yintercept=cut_points[,1], color = lc1, alpha=0.02)
plt <- plt + geom_hline(yintercept=cut_points[,2], color = lc2, alpha=0.02)
plt
```
