---
title: "Nonparametric modeling of longitudinal data"
subtitle: "CS-E5890 - Statistical Genetics and Personalised Medicine"
author: Juho Timonen
date: February 28, 2020

output:
  ioslides_presentation:
    smaller: true
    css: styles.css
header-includes:
   - \usepackage{caption}
   - \captionsetup[figure]{font=\scriptsize}
bibliography: bibliography.bib
---

# Introduction

```{r, setup, echo=FALSE, fig.height=1, message=FALSE}
#devtools::install_github('jtimonen/lgpr')
library(lgpr)
library(lme4)
library(lattice)
```


## Longitudinal studies 

- Same observational units are measured at multiple time points
- Also called *panel studies*
- Example from @belenky2003:
 
```{r, sleepstudy, echo=FALSE, fig.height=4.6}
 xyplot(Reaction ~ Days | Subject, sleepstudy,
        xlab = "Days of sleep deprivation",
        ylab = "Average reaction time (ms)", aspect = "xy")
```

## Biomedical longitudinal data

- The longitudinal study design is common in biomedicine
- Observational units are individual persons (patients)
- Makes it possible to indentify of disease predictors
- We focus on purely observational data, instead of studying the effect of some treatment


---

```{r, cars, fig.cap="A scatterplot.", echo=TRUE}
plot(cars)
```

## References